---
title: "Exploratory Analysis"
output:
  html_document:
    df_print: paged
    theme: cerulean
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r init}

source("init_data.r")
source("exploratory_functions.r")

# See what we have, here..
#foo <- convert_char_factors_to_integers(positive.churn)
```

## Data Distributions

```{r plot_distributions}
plot_distributions(train, test)
```

## Comparison of factors in Training vs Test Data

```{r plot_factor_comparisons, message=FALSE, warning=FALSE}
compare_factors(train, test)

#train.factors <- convert_to_factors(train, debug = TRUE)
#train.factors <- convert_NAs_to_level(train.factors, debug = TRUE)
#train.appetency.factors <- bin_negative_levels(train.factors, "appetency")
#train.upselling.factors <- bin_negative_levels(train.factors, "upselling")
#train.churn.factors <- bin_negative_levels(train.factors, "churn")

#summarise_levels(train.appetency.factors, "appetency")
#summarise_levels(train.upselling.factors, "upselling")
#summarise_levels(train.churn.factors, "churn")


```

## Summary of training factors are removing those levels not present in the test data
```{r plot_training_factors_after_removal, message=FALSE, warning=FALSE}
trainWithOnlyCommonLevels <- bin_levels_if_not_in_test_set(train, test, "BIN")
summarise_levels(trainWithOnlyCommonLevels, "appetency")
summarise_levels(trainWithOnlyCommonLevels, "upselling")
summarise_levels(trainWithOnlyCommonLevels, "churn")

```


## plot_distributions of +1 vs -1 categories

```{r plot_positive_negative_categories, message=FALSE, warning=FALSE}
plot_distributions(positive.appetency, negative.appetency, data1label="positive for appetency", data2label="negative for appetency")
plot_distributions(positive.upselling, negative.upselling, data1label="positive for upselling", data2label="negative for upselling")
plot_distributions(positive.churn, negative.churn, data1label="positive for churn", data2label="negative for churn")


